Server-Side Probabilistic Programming

### 2018-11-01 abstract/paper.tex
positive

\begin{abstract}
We present a view on practical probabilistic programming, as
well as an inference facility for Go programs following the
view. Probabilistic programming promises to free the
practitioner from burden of re-implementing inference paradigms
and algorithms. There are dozens of probabilistic programming
languages currently in development. Yet, integrating
probabilistic programming inference into a software system
requires greater effort than writing a custom simulation and
inference algorithm. Probabilistic programming remains, at best,
a tool for modeling and prototyping.

We believe that the probabilistic programming approach can be
made practical. However, this requires a shift from design of
languages best suited for probabilistic inference to design of
tools and libraries which make probabilistic programming in
modern general-purpose languages accessible and convenient.
Here, we first discuss the challenges in practical use of
existing probabilistic programming systems, and then suggest 
guidelines to implementation of a practical probabilistic
programming facility. We illustrate the guidelines on an
early-stage implementation of such facility for the Go
programming language. 
\end{abstract}

\begin{document}

\title{Towards Practical Probabilistic Programming}
\author{David Tolpin}
\affiliation{
    \institution{PUB+}
    \country{Israel}
}
\email{david.tolpin@gmail.com}

\maketitle

\section{Challenges in Practice of Probabilistic Programming}

Incorporating a probabilistic program, or rather a probabilistic
procedure, within a larger code body appears to be rather
straightforward: one implements the model in the probabilistic
language, fetches and preprocesses the data in the
`regular' host programming language, passes the data and the
model to an inference algorithm, and post-processes the results
in the `regular' programming language again to make algorithmic
decisions based on inference outcomes. The reality is more
complicated though, and the practitioner may face obstacles on
each step of the above routine.

\paragraph{Implementing the model} Probabilistic programming is
supposed to let the programmer use the expressive power of a
general programming language to specify probabilistic models.
However, the probabilistic language is often either a small
language lacking advanced control and data manipulation
capabilities (Anglican~\cite{TMY+16}, Stan~\cite{Stan17},
WebPPL~\cite{GS15}), or an API to the representation of the
probabilistic model (Edward~\cite{THS+17}, Figaro~\cite{P09}).
When the probabilistic language is the same or as rich as the
host language, its use is still constrained by available data
structures (Pyro~\cite{Pyro18}), or, the host language is 
uncommon as a general-purpose language (Turing~\cite{GXG18}).
Often, the probabilistic language is interpreted, making
execution of the model too slow.

\paragraph{Passing the data} The case when all observations are
handed over to inference in bulk, and a short summary of
inference results returned, can mostly be implemented
efficiently. This use case is well addressed by most
environments even when the host language differs from the
probabilistic language. A difficulty arises when
the data arrives incrementally, and inference outcomes are
processed in any-time manner. Data has to be marshalled when the
host and probabilistic language platforms differ (e.g.
Stan), or transferred between data structures used in
the host language and the data structures required by the
probabilistic language (e.g. Pyro).

\paragraph{Simulation-inference dichotomy} Many probabilistic
models follow a design pattern of simulation-inference: a
significant part of the model is a simulator, running an
algorithm with fixed parameter; the optimal parameters, or their
distribution, are to be inferred. The simulator may be quite
complicated, however in addition to the apparent difficulty of
implementing the simulator in a less expressive language,  the
simulator has to be implemented twice: first in the
probabilistic language for inference, and then in the host
environment to make use of the inferred parameters. This not
just involves extra effort but is also prone to introducing
programming mistakes, when the two versions of the simulator
behave differently for the same parameter values. One possible
solution is to treat the simulator as a black box~\cite{LBW17},
however better inference algorithms can be used if the structure
of the simulator is revealed to the probabilistic programming
facility, e.g. through the use of automatic differentiation.

\paragraph{Deployment} A software system composed of
heterogeneous languages and environments is harder to integrate
and deploy. Many software solutions are monolithic, comprising
of just a single programming language and with a conservative
set of dependencies. Adding a probabilistic programming system,
often depending on a different runtime and libraries, makes
integration and deployment more complicated.

\section{Guidelines to a Practical Probabilistic Programming
Facility}

Based on the experience of implementing  probabilistic
programming languages, as well as developing and deploying
solutions using different probabilistic environments, we suggest
guidelines to implementation of a probabilistic programming
facility. We believe that these guidelines, if followed, will
lead to eventual wide adoption of probabilistic programming in
software engineering.

@@

Following the guidelines, we have implemented a probabilistic
programming facility for the Go programming language,
\texttt{infergo} (\url{http://infergo.ml/}). We have chosen Go
because Go is a relatively small programming language with
efficient implementation, which has recently become quite
popular for computation-intensive server-side programming. At
the time of writing, the implementation is still at an early
stage. However, this facility is already used in production
environment for inference of mission-critical algorithm
parameters.

@@

While we enjoy the ease and efficiency of
\texttt{infergo}-based inference in our production environment,
we view \texttt{infergo} as a reference implementation rather than
as the only solution, and encourage the community to equip other
general-purpose languages with similar inference facilities,
which, we believe, will make wide adoption of probabilistic
programming more likely.
be

>>>>

\title{Server-Side Probabilistic Programming}
\author{David Tolpin}
\affiliation{
    \institution{PUB+}
    \country{Israel}
}
\email{david.tolpin@gmail.com}

\begin{abstract}
	We share our experiences from integrating probabilistic
	programming into a server-side software system and
	present our ongoing work on probabilistic programming
	facility for Go, a modern programming language of choice for
	server-side software development. We demonstrate how a
	lightweight efficient probabilistic programming facility can
	be added to an existing programming language.  Server-side
	application of probabilistic programming poses challenges
	for a probabilistic programming system. We discuss the
	challenges and our experience in overcoming them, and
	suggest guidelines which can help in a wider adoption of
	probabilistic programming in server-side software systems.
\end{abstract}

\begin{document}
\maketitle

\section{Challenges of Server-Side Probabilistic Programming}

Incorporating a probabilistic program, or rather a probabilistic
procedure, within a larger code body appears to be rather
straightforward: one implements the model in the probabilistic
programming language, fetches and preprocesses the data in the
host programming language, passes the data and the model to an
inference algorithm, and post-processes the results in the
host programming language again to make algorithmic
decisions based on inference outcomes. However, complex
server-side software systems make integration of probabilistic
inference challenging.

\paragraph{Simulation-inference dichotomy} Probabilistic models
often follow the design pattern of simulation-inference: a
significant part of the model is a simulator, running an
algorithm with fixed parameters; the optimal parameters, or
their distribution, are to be inferred. The inferred parameters
are then used by the software system to execute the simulation
independently of inference for forecasting and decision making.

This pattern suggest re-use of the simulator: instead of
implementing the simulator twice, in the probabilistic model and
in the host system, the same code can serve both purposes.
However to achieve this, the host language must coincide with
the implementation language of the probabilistic model, on one
hand, and allow a computationally efficient implementation of
the simulation, on the other hand. Some probabilistic systems
(Figaro~\cite{P09}, Anglican~\cite{TMY+16}, Turing~\cite{GXG18})
are built with tight integration with the host environment in
mind; more often than not though the probabilistic code is
not trivial to re-use.

\paragraph{Data interface} In a server-side application data
for inference comes from a variety of sources: network,
databases, distributed file systems, and in many different
formats. Efficient inference depends on fast data access and
updating. Libraries for data access and manipulation are
available in the host environment. While the host environment
can be used as a proxy retrieving and transforming the data,
such as in the case of Stan~\cite{Stan17} integrations,
sometimes direct access from the probabilistic code is the
preferred option, for example when the data is streamed or
retrieved conditionally.

\paragraph{Integration and deployment} Deployment of server-side
software systems is a delicate process involving automatic
builds and maintenance of dependencies. Adding  a component
which possibly introduces additional software dependencies or
even a separate runtime complicates deployment. Minimizing the
burden of probabilistic programming on the integration
and deployment process should be a major consideration in
design or selection of probabilistic programming tools.
Probabilistic programming systems which are implemented or 
provide an interface in a popular programming language, e.g.
Python (Edward~\cite{THS+17}, Pyro~\cite{Pyro18}) are easier
to integrate and deploy, however the smaller the footprint
of a probabilistic system, the easier is the adoption.

Based on the experience of developing and deploying solutions
using different probabilistic environments, we came up with
guidelines to implementation of a probabilistic programming
facility for server-side applications. We believe that these
guidelines, when followed, help easier integration of
probabilistic programming inference into large-scale server-side
software systems.

@@

In line with the guidelines, we have implemented a probabilistic
programming facility for the Go programming language,
\texttt{infergo} (\url{http://infergo.ml/}). We have chosen Go
because Go is a small but expressive programming language with
efficient implementation, which has recently become quite
popular for computation-intensive server-side programming. At
the time of writing, the implementation is still at an early
stage. However, this facility is already used in production
environment for inference of mission-critical algorithm
parameters.

@@

The Go programming language and development environment offer
capabilities which made implementation of \texttt{infergo}
affordable.

\begin{enumerate}
	\item The Go parser and abstract syntax tree serializer are a
		part of the standard library. Reading, transforming, and
		writing Go source code is straightforward and effortless.
	\item Type inference (or \textit{type checking} as it is
		called in the Go ecosystem), also provided in the
		standard library, augments parsing and allows to
		selectively apply transformation-based automatic
		differentiation  based on static expression types. 
	\item Go compiles and runs fast. Fast compilation and
		execution speeds allow to use the same facility for both
		exploratory design of probabilistic models and for
		inference in production environment.
	\item Go offers efficient parallel execution as a
		first-class feature, via so-called \textit{goroutines}.
		Goroutines streamline implementation of sampling-based
		inference algorithms. Sample generators and consumers
		are run in parallel, communicating through channels. 
		Inference is easy to parallelize in order to exploit
		hardware multi-processing, and samples are retrieved
		lazily for postprocessing. 
\end{enumerate}

A lightweight probabilistic programming facility similar to
\texttt{infergo} can be added to most modern general-purpose
programming languages, in particular those used in implementing
large-scale software systems, making probabilistic
programming inference more accessible in server-side
applications.
